---
title: "AllelicImbalance Vignette"
author: "Jesper R. Gådin and Lasse Folkersen"
date: "`r Sys.Date()`"
output:
  BiocStyle::pdf_document:
    toc: true
vignette: >
  %\VignetteIndexEntry{AllelicImbalance Vignette}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---
```{r style, echo = FALSE, results = 'asis'}
BiocStyle::markdown()
```
#Introduction
This AllelicImbalance package contains functions for investigating
allelic imbalance effects in RNA-seq data. Maternal and paternal
alleles could be expected to show identical transcription rate, resulting in a
50%-50% mix of maternal and paternal mRNA in a sample. However, this turns out
to sometimes not be the case. The most extreme example is the X-chromosome
inactivation in females, but many autosomal genes also have deviations from
identical transcription rate. The causes of this are not always known, but one
likely cause is the difference in DNA, namely heterozygous SNPs, affecting
enhancers, promoter regions, splicing and stability. Identifying this allelic
imbalance is therefore of interest to the characterization of the genome and the
aim of the AllelicImbalance package is to facilitate this.

Load AllelicImbalance
```{r echo = TRUE, eval = TRUE, message = FALSE}
library(AllelicImbalance)
```

# ASEset
The ASEset object is the central class of objects in the AllelicImbalance package. The ASEset object has the SummarizedExperiment as parent class, and all functions you can apply on this class you can also apply on an ASEset.

## Simple example of building an ASEset object
In this section we will walk through the various ways an ASEset
object can be created. Although the preprocessing of RNA-seq data is not the
primary focus of this package, it is a necessary step before analysis. There
exists several different methods for obtaining a bam file, and this section
should just be considered an example. For further details we refer
to the web-pages of tophat, bowtie, bwa and samtools found in the links section at the end of this document.

```
wget ftp://ftp.sra.ebi.ac.uk/vol1/fastq/ERR009/ERR009135/*
bowtie -q --best --threads 5 --sam hg19 +
>      -1 ERR009135_1.fastq.gz -2 ERR009135_2.fastq.gz "ERR009135.sam"
samtools view -S -b ERR009135.sam > ERR009135.bam
```

In the above code one paired-end RNA sequencing sample is downloaded and
aligned to the human genome, then converted to bam using samtools. The
resulting bam files can be the direct input to the AllelicImbalance package. Other
aligners can be used as well, as long as bam files are provided as input. The
example code following illustrates how to use the import mechanism on a
chromosome 17-located subset of 20 RNA-seq experiments of HapMap samples. The
output is an ASEset object containing allele counts for all
heterozygote coding SNPs in the region.

```{r echo = TRUE, eval = TRUE}
searchArea <- GRanges(seqnames = c("17"), ranges = IRanges(79478301, 79478361))
pathToFiles <- system.file("extdata/ERP000101_subset", package = "AllelicImbalance")
reads <- impBamGAL(pathToFiles, searchArea, verbose = FALSE)
heterozygotePositions <- scanForHeterozygotes(reads, verbose = FALSE)
countList <- getAlleleCounts(reads, heterozygotePositions, verbose = FALSE)
a.simple <- ASEsetFromCountList(heterozygotePositions, countList)
a.simple
```

## Building an ASEset object using Bcf or Vcf files
If more than a few genes and a few samples are analyzed we recommend
that a SNP-call is instead made using the samtools mpileup function (see
links section). The scanForHeterozygotes function is merely a simple 
SNP-caller and it is not as computationally optimized as e.g. mpileup. In this bash
code we download reference sequence for chromosome 17 and show how to generate
mpileup calls on one of the HapMap samples that were described above.

```
wget ftp://hgdownload.cse.ucsc.edu/goldenPath/hg19/chromosomes/chr17.fa.gz
samtools mpileup -uf hg19.fa ERR009135.bam | bcftools view -bvcg - > ERR009135.bcf
```

### Vcf-version
Simply use bcftools to get back to the vcf format

```
bcftools view > ERR009135.vcf
```

Use the readVcf function from the VariantAnnotation package

```{r echo = TRUE, eval = TRUE, warning = FALSE}
pathToVcf <- paste(pathToFiles,"/ERP000101.vcf",sep="")
VCF <- readVcf(pathToVcf,"hg19") 
gr <- granges(VCF) 

#only use bi-allelic positions
gr.filt <- gr[width(mcols(gr)[,"REF"]) == 1 | 
				unlist(lapply(mcols(gr)[,"ALT"],width))==1]

countList <- getAlleleCounts(reads, gr.filt, verbose=FALSE) 
a.vcf <- ASEsetFromCountList(rowData = gr, countList)
```

### Bcf-version
Samtools mpileup generates by default a Vcf file which contains SNP and short INDEL 
positions. Piping the output to bcftools we get its binary equivalent (Bcf), which 
takes less space and can be queried more effective. With the Bcf files the process 
of generating an ASEset object starts with a call to the impBcfGR 
function instead. This function will import the Bcf file containing all SNP calls 
that were generated with the samtools mpileup function.

```{r echo = TRUE, eval = TRUE}
BcfGR <- impBcfGR(pathToFiles,searchArea,verbose=FALSE)
countListBcf <- getAlleleCounts(reads, BcfGR,verbose=FALSE)
a.bcf <- ASEsetFromCountList(BcfGR, countListBcf)
```

## Using strand information
Many RNA-seq experiments do not yield useful information on the strand from
which a given read was made. This is because they involve a step in which a
double-stranded cDNA is created without tracking strand-information. Some
RNA-seq setups do however give this information and in those cases it is important to
keep track of strand in the ASE-experiment. The example data from above is
from an experiment which created double-stranded cDNA before labelling and so the '+'
and '-' information in it is arbitrary. However, if we assume that the
information has strand information, then the correct procedure is as follows:


```{r echo = TRUE, eval = TRUE}
plus <- getAlleleCounts(reads, heterozygotePositions, strand="+",verbose=F) 
minus <- getAlleleCounts(reads, heterozygotePositions, strand="-",verbose=F)

a.stranded <-
ASEsetFromCountList(
heterozygotePositions,
countListPlus=plus,
countListMinus=minus
)
a.stranded
```
The main effect of doing this, is in the plotting functions which will separate
reads from different strands if they are specified as done here. It is
important, however, to make sure that the imported RNA-seq experiment does in
fact have proper labeling and tracking of strand information before proceeding
with this method.

## Two useful helper functions
At this stage it is worth highlighting two useful helper functions that both
uses existing BioC annotation objects. One is the
getAreaFromGeneNames which quickly retrieves the above mentioned
searchArea when given just genesymbols as input, and relies on
 `r Biocannopkg("org.Hs.eg.db")`. The other other is
the getSnpIdFromLocation function which attempts to rename
location-based SNP names to established rs-IDs in case they exist. These
functions work as follows:

```{r echo = TRUE, eval = TRUE, message = FALSE, results = 'hide'}
#Getting searchArea from genesymbol
library(org.Hs.eg.db )
searchArea<-getAreaFromGeneNames("ACTG1",org.Hs.eg.db)

#Getting rs-IDs
library(SNPlocs.Hsapiens.dbSNP.20120608)
#seqlevels(a.simple) <- "chr17"
gr <- rowRanges(a.simple)
updatedGRanges<-getSnpIdFromLocation(gr, SNPlocs.Hsapiens.dbSNP.20120608)
rowRanges(a.simple)<-updatedGRanges
```

## Adding phenotype data
Typically an RNA-seq experiment will include additional information about each
sample. It is an advantage to include this information when creating an ASEset
because it can be used for subsequent highlights or subsetting in plotting and
analysis functions.

```{r echo=TRUE, eval=TRUE}
#simulate phenotype data
pdata <- DataFrame(
	Treatment=sample(c("ChIP", "Input"),length(reads),replace=TRUE),
	Gender=sample(c("male", "female"),length(reads),replace=TRUE), 
	row.names=paste("individual",1:length(reads),sep=""))

#make new ASEset with pdata
a.new <- ASEsetFromCountList(
		heterozygotePositions,
		countList,
		colData=pdata)

#add to existing object
colData(a.simple) <- pdata
```

## Adding phase information
For some functionality phase information is necessary. Phasing can be obtained from many external sources e.g. samtools. The phase information is often present in VCF-files. And the lines below show how to access that information and transfer it to an ASEset. The ASEset follows the VCF-conventions on how to describe the phase, i.e. each patients phase will be described by the established notation of the form "1|0","1|1" or "1/0". There the left number is the description of the maternal allele and the right number is the description of the paternal allele. If it is 0 the allele is the same as the reference allele, and if it is 1 it is the alternative allele. For "|" the phase is known and for "/" the phase is not known. Note, that in the AllelicImbalance package only bi-allelic expression is allowed.

```{r echo=TRUE, eval=TRUE}
#manually add the phase information
#example phase matrix 
set.seed(1)
rows <-nrow(a.simple)
cols <- ncol(a.simple)
p1 <- matrix(sample(c(1,0),replace=TRUE, size=rows*cols), nrow=rows, ncol=cols)
p2 <- matrix(sample(c(1,0),replace=TRUE, size=rows*cols), nrow=rows, ncol=cols)
phase.mat <- matrix(paste(p1,sample(c("|","|","/"), size=rows*cols, replace=TRUE), p2, sep=""),
	nrow=rows, ncol=cols)
phase(a.simple) <- phase.mat 
```

```{r echo=TRUE, eval=TRUE}
#use the phase information from a Vcf file
pathToVcf <- system.file("extdata/ERP000101_subset/ERP000101.vcf", package = "AllelicImbalance")
p <- readGT(pathToVcf)
#The example Vcf file contains only 19 out of our 20 samples
#So we have to subset and order
a.subset <- a.simple[,colnames(a.simple) %in% colnames(p)]
p.subset <- p[, colnames(p) %in% colnames(a.subset)]
p.ordered <- p.subset[ , match(colnames(a.subset), colnames(p.subset))]
```

##Adding reference and alternative allele information
The reference and alternative alleles are stored in the meta-columns
and can be accessed and set through the mcols() function. All functions
that require reference or alternative allele will check
in the meta-columns "ref" or "alt" for that information.

```{r echo=TRUE, eval=TRUE}
#from simulated data
mcols(a.simple)[,"ref"] <- c("T","C","G")

#from reference genome
data(ASEset.sim)
fasta <- system.file('extdata/hg19.chr17.subset.fa', package='AllelicImbalance')
a.ref <- refAllele(ASEset.sim,fasta=fasta) 

#infer and set alternative allele
alt <- inferAltAllele(a.simple)
mcols(a.simple)[,"alt"] <- alt
```

#Tests
## Statistical analysis of an ASEset object
One of the simplest statistical test for use in allelic imbalance analysis is
the chi-square test. This test assumes that the uncertainty of ASE is represented by a normal distribution around an expected mean (i.e 0.5 for equal expression). A significant result suggests an ASE event. Every strand is tested independently. 

```{r echo=TRUE, eval=TRUE}
#use a subset for tests
a2 <- a.stranded[,5:10]
#binomial test
binom.test(a2,"+")
#chi-square test
chisq.test(a2,"-")
```

# Standard Graphics
## Plotting of an ASEset object
The barplot function for ASEset objects
plots the read count of each allele in each sample. This is useful for getting a very
detailed view of individual SNPs in few samples. As can be seen below, four
samples from the HapMap data contains a strong imbalance at the chr17:79478331
position on the plus strand. By default the p-value is calculated by a
chi-square test. To use other test results the arguments testValue and
testValue2 can be used. When the counts for one allele are below 5 for
one allele the chi-square test returns NA. This is why there is no P-value above the first bar in the example below.

```{r echo=TRUE, eval=TRUE}
barplot(a.stranded[1], strand="+", xlab.text="", legend.interspace=2)

#set own computed p-values
btp <- binom.test(a.stranded[1],"+")
barplot(a.stranded[1], strand="+", testValue=btp, xlab.text="", legend.interspace=2)
```

Another example of plotting that is useful is the one invoked with the plotting
type argument "fraction". This plotting mechanism is
useful to illustrate more SNPs and more samples in less space than the standard
plot. As can be seen here several other samples are not heterozygote at the
chr17:79478331 location.

```{r echo=TRUE, eval=TRUE}
barplot(a.simple[1], type="fraction", cex.main = 0.7)
```

A typical question would be to ask why certain heterozygote samples have allele
specific expression. The argument sampleColour argument allows for
different highligts such as illustrated here below for gender. This could also
be used to highlight based on genotype of proximal non-coding SNPs if available.

```{r echo=TRUE, eval=TRUE}
sampleColour<-rep("palevioletred",ncol(a.simple))
sampleColour[colData(a.simple)[,"Gender"]%in%"male"] <- "blue"
barplot(a.simple[1], type="fraction", sampleColour=sampleColour, cex.main = 0.7)
```

## Plot with annotation
It is often of interest to combine the RNA
sequencing data with genomic annotation information from online databases. For
this purpose there is a function to extract variant specific annotation such as
gene, exon, transcript and CDS.

```{r echo=TRUE, eval=TRUE, message=FALSE}
library(org.Hs.eg.db)
barplot(a.simple[1],OrgDb=org.Hs.eg.db,
			cex.main = 0.7,
			xlab.text="",
			ypos.annotation = 0.08,
			annotation.interspace=1.3,
			legend.interspace=1.5
			)
```

## locationplot
Finally a given gene or set of proximal genes will often have several SNPs close
to each other. It is of interest to investigate all of them together, in
connection with annotation information. This can be done using the
locationplot function. This function in its simplest form just plot
all the SNPs in an ASEset distributed by genomic location. Additionally it
contains methods for including gene-map information through the arguments
OrgDb and TxDb.

```{r echo=TRUE, eval=TRUE, message=FALSE}
library(TxDb.Hsapiens.UCSC.hg19.knownGene)
#using count type
locationplot(a.stranded, strand="-", type="count", cex.main=0.7, cex.legend=0.4)
#use annotation
a.stranded.sorted <- sort(a.stranded, decreasing=FALSE)
locationplot(a.stranded.sorted, strand="-", OrgDb=org.Hs.eg.db,TxDb=TxDb.Hsapiens.UCSC.hg19.knownGene)
```

#Grid Graphics
To make use of the extra graphical flexibility that comes with grid graphics, 
the AllelicImbalance package has now functions that can be integrated in that
environment, visit this [link](https://www.stat.auckland.ac.nz/~paul/grid/grid.html) for more information. Another good reason for grid graphics is the Gviz package, which
uses tracks to visualize genomic regions in a smart way. Below are examples
for the bar- and locationplot versions using the grid, having an extra "g" as prefix in the function name.  

##lbarplot
The standard barplot function offers a lot 

```{r echo=TRUE, eval=TRUE, message=FALSE}
#access example ASEset data
data(ASEset)

#lbarplots with type "count" 
lbarplot(ASEset, type="count")

# lbarplots with type "fraction" 
lbarplot(ASEset, type="fraction")
```

##glocationplot
The "g" in glocationplot comes from "Gviz barplot". It is a wrapper to quickly 
get an overview of ASE in a region.

```{r echo=TRUE, eval=TRUE, message=FALSE}
#remember to set the genome
genome(ASEset) <- "hg19"

glocationplot(ASEset,strand='+')

#for ASEsets with fewer SNPs the 'count' type plot is useful 
glocationplot(ASEset[,1:5], type="count")
```

##Custom locationplots
More flexibility and functionality from the Gviz package is accessed 
if the tracks are constructed separately.

```{r echo=TRUE, eval=TRUE, message=FALSE}
#subset ASEset and reads
x <- ASEset[1:2]
r <- reads[1:2]
seqlevels(r, force=TRUE) <- seqlevels(x)

GR <- GRanges(seqnames=seqlevels(x),
 		ranges=IRanges(start=min(start(x)),end=max(end(x))),
 		strand='+', genome=genome(x))

deTrack <- ASEDAnnotationTrack(x, GR=GR, type='fraction',strand='+')
covTracks <- CoverageDataTrack(x, BamList=r, strand='+') 
lst <- c(deTrack,covTracks)
sizes <- c(0.5,rep(0.5/length(covTracks),length(covTracks)))

plotTracks(lst, from=min(start(x)), to=max(end(x)), sizes=sizes, col.line = NULL, showId = FALSE, main = 'main', cex.main = 1, title.width = 1, type = 'histogram')
```

##Top allele criteria
The barplot with type='function' can be visualized according to three different allele sorting criteria. The default behaviour is just to shown the allele with highest overall abundance in the upper half of the plot. This works well for most single SNP investigations. For more complex situations, however, it can be essential to keep track of phase information. This is done either through the reference allele sorting function, or even better, through consistently showing the maternal allele on top. When phase is know, this is essential to compare effect-directions of different coding SNPs

```{r echo=TRUE, eval=TRUE, message=FALSE}

#load data
data(ASEset)

#using reference allele information
mcols(ASEset)[,"alt"] <- inferAltAllele(ASEset)
barplot(ASEset[1], type="fraction", strand="+", xlab.text="", top.allele.criteria="ref")

#using phase information
phase(ASEset) <- phase.mat 
barplot(ASEset[1], type="fraction", strand="+", xlab.text="", top.allele.criteria="phase")
```

# Summary functions
The regionSummary function can be used to investigate if there is a consistent imbalance in the same direction over a region (e.g. a transcript). 

```{r echo=TRUE, eval=TRUE, message=FALSE}
# in this example every snp is on its own exon
txGR <- granges(ASEset)
rs <- regionSummary(ASEset, txGR)
head(rs)
```


# Conclusion
In conclusion we hope that you will find this package useful in the
investigation of the genetics of RNA-seq experiments. The various import
functions should assist in the task of actually retrieving allele counts for
specific nucleotide positions from all RNA-seq reads, including the non-trivial
cases of intron-spanning reads. Likewise, the statistical analysis and plotting
functions should be helpful in discovering any allele specific expression
patterns that might be found in your data.

# Links

Bowtie
[link](http://bowtie-bio.sourceforge.net)

BWA
[link](http://bio-bwa.sourceforge.net/)

Samtools
[link](http://samtools.sourceforge.net/)

Samtools pileup
[link](http://samtools.sourceforge.net/mpileup.shtml)


> "Who is is the master if nature provides the techniques to study itself"


# Session Info
```{r echo=TRUE, eval=TRUE}
sessionInfo()
```
